\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-03-13 jasa/paper-jasa.tex
new figures
minor text changes

In the case of composite hypotheses $\theta \in \Theta_0$ against an alternative
$\theta \in \Theta_1$ (such that $\Theta_0 \cap \Theta_1 = \emptyset$ and $\Theta_0 \cup \Theta_1 = \Theta$), the
generalized likelihood ratio test, also known as the profile likelihood ratio
test, is commonly used
\begin{equation}\label{eqn:generalized-lr}
\Lambda({\cal D}; \Theta_0, \Theta) =  \frac{ \sup_{\theta \in \Theta_0} p({\cal D} | \theta)}{ \sup_{\theta \in \Theta} p({\cal D} | \theta)} \; .
\end{equation}
This generalized likelihood ratio can be used both for hypothesis tests in the
presence of nuisance parameters or to create confidence intervals with or
without nuisance parameters.  Often, the parameter vector is broken into two
components $\theta=(\mu,\nu)$, where the $\mu$ components are considered
parameters of interest while the $\nu$ components are considered nuisance
parameters. In that case $\Theta_0$ corresponds to all values of $\nu$ with
$\mu$ fixed.

@@

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig1a.pdf}
        \caption{$p(\mathbf{x}|\gamma=0.05)$ and $p(\mathbf{x}|\gamma=0)$}
        \label{fig:1a}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig1b.pdf}
        \caption{Linear model}
        \label{fig:1b}
    \end{subfigure}

@@

@@

As shown in the figures, qualitatively different results are obtained depending
the classification algorithm embedded within the approximation of the ratio. The
neural network-based approximation appears to work best, due to the capacity of
the classifier to model a smooth decision function $s(\mathbf{x})$ which is
very close to the optimal decision function. The random forest yields a decision function
which is piece-wise constant, therefore non-continuous and not strictly
monotonic with the ratio in the finite sample regime. On this problem, the
random forest still approximates the ratio correctly, but more variance can be
observed. Finally, the linear model produces a decision function that is
smooth and continuous but only piece-wise monotonic with the ratio. Due to the symmetry of
the ratio around $\mathbf{x}=1$, this still results in an accurate
approximation, but poorer results should be expected on more complex problems.
While not directly apparent, calibrating $\hat s(\mathbf{x})$ to construct $\hat p(\hat
s(\mathbf{x}))$ is key to obtain accurate results. Standard
histograms with fixed bins have been used here for illustrative purposes, but we
anticipate that the embedded density estimation algorithm should be carefully
chosen and tuned, on a case by case basis. Overall, this first example confirms
that even if the classifier does a poor job at modeling the optimal decision
function, good approximations of the likelihood ratio can still be obtained
with the proposed method, by calibrating $s(\mathbf{x})$ and/or by decomposing
the mixture when possible.

@@

Given observed data from Figure~\ref{fig:1a}, let us now consider the composite hypothesis testing setting, by comparing
$H_0: \gamma = 0.05$ versus the alternative $H_1: \gamma \neq 0.05$. That is, we want to evaluate
\begin{equation}
\Lambda({\cal D}; \Theta_0 = \{0.05\}, \Theta=[0, 1]) =  \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} \; .
\end{equation}
By exploiting the fact that
\begin{equation}
\log \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} = \log \frac{ p({\cal D} | \gamma = 0.05) }{ p({\cal D} | \gamma = 0 ) } - \log \frac{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma) }{  p({\cal D} | \gamma = 0 ) },
\end{equation}
the generalized likelihood ratio statistic can be computed by evaluating
both terms reusing the decomposed neural network-based approximation from Figure~\ref{fig:1c}.
In particular, the maximum likelihood estimator (MLE) in the second term
can be found as outlined in
Section~\ref{sec:generalized-likelihood-ratio}.
Figure~\ref{fig:2a} shows the empirical distribution of the resulting MLEs,
compared to the distribution of the true MLEs. It clearly suggests that
the proposed inference procedure yields in this case an unbiased estimator
($\mathbb{E} [\hat \gamma] = 0.0501$, as empirically estimated over $1000$
trials). Note that had the underlying ratio not been accurate enough because of
poor training or poor calibration, the two distributions would not be exactly
overlapping, indicating bias. From this, the initial $\Lambda$ test statistic can finally
be evaluated to determine in this case that the null should not be rejected.  Figure \ref{fig:2b} also confirms
that $-2 \log \Lambda$, as approximated with our method, is asymptotically $\chi^2_1$ distributed,
as should be the true statistic when the null is true.

@@

Given observations ${\cal D}$ represented in Figure~\ref{fig:4},
our goal is to infer the nominal values $\alpha=1$ and $\beta=-1$ that
were used for generating this data.
We use the objective function
\begin{equation}\label{eqn:obj-ex2}
    -2 \log \frac{p({\cal D} | \alpha, \beta)}{ \sup_{\alpha,\beta} p({\cal D} | \alpha, \beta) }
\end{equation}
that we approximate by exploiting the fact that
\begin{equation}\label{eqn:obj-ex2-expanded}
\log \frac{p({\cal D} | \alpha, \beta)}{ \sup_{\alpha,\beta} p({\cal D} | \alpha, \beta) } = \log \frac{p({\cal D} | \alpha, \beta)}{ p({\cal D} | \alpha=0, \beta=0) } - \log \frac{ \sup_{\alpha,\beta} p({\cal D} | \alpha, \beta) }{ p({\cal D} | \alpha=0, \beta=0)}.
\end{equation}
Following recommendations from Section~\ref{sec:param-clf} to reduce
computational requirements, both terms in the right hand side of Eqn.~\ref{eqn:obj-ex2-expanded} are evaluated using a single
parameterized approximated  likelihood ratio, in this case based on a parameterized
2-layer neural network where the alternate is fixed to $\alpha=0, \beta=0$.
Since the generative model is not expensive, the classifier output is calibrated
on-the-fly with histograms for  every candidate parameter pair $(\alpha,\beta)$.
Minimizing the constructed approximation of the objective with
Bayesian optimization~\citep{brochu2010tutorial} eventually leads to the
estimated posterior mean shown in Figure~\ref{fig:5c}, which is close to the
true objective illustrated in Figure~\ref{fig:5a}. After $50$ iterations, the
estimated minimizer is found at $(\hat \alpha=1.008, \hat \beta=-1.004)$. For
comparison, Figure~\ref{fig:5b} shows the objective value as directly
approximated by the proposed method for a fixed grid of values. Since $\hat s(\mathbf{x;\alpha,\beta})$ is calibrated on-the-fly,
roughness in the contours can be observed, in particular due to non-smooth changes in the resulting histograms as $\alpha$ and $\beta$ vary.
As discussed in Section~\ref{sec:param-calibration}, parameterized calibration
is a solution to this issue, but is left for now as an area for future work.
Overall, this example further illustrates and confirms the ability
of the proposed method for inference, even in the case of multi-dimensional data where
direct approximations $p(\mathbf{x}|\theta_0)$ and $p(\mathbf{x}|\theta_1)$
are often difficult to construct.

>>>>

In the case of composite hypotheses $\theta \in \Theta_0$ against an alternative
$\theta \in \Theta_1$ (such that $\Theta_0 \cap \Theta_1 = \emptyset$ and $\Theta_0 \cup \Theta_1 = \Theta$), the
generalized likelihood ratio test, also known as the profile likelihood ratio
test, is commonly used
\begin{equation}\label{eqn:generalized-lr}
\Lambda(\Theta_0) =  \frac{ \sup_{\theta \in \Theta_0} p({\cal D} | \theta)}{ \sup_{\theta \in \Theta} p({\cal D} | \theta)} \; .
\end{equation}
%\begin{equation}\label{eqn:generalized-lr}
%\Lambda({\cal D}; \Theta_0, \Theta) =  \frac{ \sup_{\theta \in \Theta_0} p({\cal D} | \theta)}{ \sup_{\theta \in \Theta} p({\cal D} | \theta)} \; .
%\end{equation}
This generalized likelihood ratio can be used both for hypothesis tests in the
presence of nuisance parameters or to create confidence intervals with or
without nuisance parameters.  Often, the parameter vector is broken into two
components $\theta=(\mu,\nu)$, where the $\mu$ components are considered
parameters of interest while the $\nu$ components are considered nuisance
parameters. In that case $\Theta_0$ corresponds to all values of $\nu$ with
$\mu$ fixed.

@@

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig1a.pdf}
        \caption{$p(\mathbf{x}|\gamma=0.05)$ and $p(\mathbf{x}|\gamma=0)$}
        \label{fig:1a}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
      %(or a blank line to force the subfigure onto a new line)
%    \begin{subfigure}[b]{0.45\textwidth}
%        \includegraphics[width=\textwidth]{figures/fig1b.pdf}
%        \caption{Linear model}
%        \label{fig:1b}
%    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/p_sij.png}
        \caption{Pairwise classifier score distributions}
        \label{fig:p_sij}
    \end{subfigure}

@@

%    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
%      %(or a blank line to force the subfigure onto a new line)
%    \begin{subfigure}[b]{0.3\textwidth}
%        \includegraphics[width=\textwidth]{figures/fig_ratio_dist.png}
%        \caption{Likelihood ratio distributions}
%        \label{fig:decomposed_ratio}
%    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig_ratio_dist.png}
        \caption{Likelihood ratio distributions}
        \label{fig:decomposed_ratio}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/likelihood.png}
        \caption{log-likelihood curve }
        \label{fig:likelihood}
    \end{subfigure}

@@

As shown in the figures, qualitatively different results are obtained depending
the classification algorithm embedded within the approximation of the ratio. The
neural network-based approximation appears to work best, due to the capacity of
the classifier to model a smooth decision function $s(\mathbf{x})$ which is
very close to the optimal decision function. The random forest yields a decision function
which is piece-wise constant, therefore non-continuous and not strictly
monotonic with the ratio in the finite sample regime. On this problem, the
random forest still approximates the ratio correctly, but more variance can be
observed. Finally, the linear model produces a decision function that is
smooth and continuous but only piece-wise monotonic with the ratio. Due to the symmetry of
the ratio around $\mathbf{x}=1$, this still results in an accurate
approximation, but poorer results should be expected on more complex problems.
While not directly apparent, calibrating $\hat s(\mathbf{x})$ to construct $\hat p(\hat
s(\mathbf{x}))$ is key to obtain accurate results. Standard
histograms with fixed bins have been used here for illustrative purposes, but we
anticipate that the embedded density estimation algorithm should be carefully
chosen and tuned, on a case by case basis. Overall, this first example confirms
that even if the classifier does a poor job at modeling the optimal decision
function, good approximations of the likelihood ratio can still be obtained
with the proposed method, by calibrating $s(\mathbf{x})$ and/or by decomposing
the mixture when possible.
%
%
%\begin{figure}
%    \centering
%    \begin{subfigure}[b]{0.3\textwidth}
%        \includegraphics[width=\textwidth]{figures/p_sij.png}
%        \caption{Pairwise classifier score distributions}
%        \label{fig:p_sij}
%    \end{subfigure}
%    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
%      %(or a blank line to force the subfigure onto a new line)
%    \begin{subfigure}[b]{0.3\textwidth}
%        \includegraphics[width=\textwidth]{figures/fig_ratio_dist.png}
%        \caption{Likelihood ratio distributions}
%        \label{fig:decomposed_ratio}
%    \end{subfigure}
%    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
%      %(or a blank line to force the subfigure onto a new line)
%    \begin{subfigure}[b]{0.3\textwidth}
%        \includegraphics[width=\textwidth]{figures/likelihood.png}
%        \caption{log-likelihood curve }
%        \label{fig:likelihood}
%    \end{subfigure}
%
%    \caption{Approximation of the log-likelihood ratio
%             $\log r(\mathbf{x};\gamma=0.05,\gamma=0) = \log \frac{p(\mathbf{x}|\gamma=0.05)}{p(\mathbf{x}|\gamma=0)}$
%             with calibrated classifiers.}
%\end{figure}
%
%%%%%%%%%%%%%

@@

Given observed data from Figure~\ref{fig:1a}, let us now consider the composite hypothesis testing setting, by comparing
$H_0: \gamma = 0.05$ versus the alternative $H_1: \gamma \neq 0.05$. That is, we want to evaluate
\begin{equation}
\Lambda( \gamma = 0.05) =  \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} \; .
\end{equation}
%\begin{equation}
%\Lambda({\cal D}; \Theta_0 = \{0.05\}, \Theta=[0, 1]) =  \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} \; .
%\end{equation}
By exploiting the fact that
\begin{equation}
\log \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} = \log \frac{ p({\cal D} | \gamma = 0.05) }{ p({\cal D} | \gamma = 0 ) } - \log \frac{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma) }{  p({\cal D} | \gamma = 0 ) },
\end{equation}
the generalized likelihood ratio statistic can be computed by evaluating
both terms reusing the decomposed neural network-based approximation from Figure~\ref{fig:1c}.
In particular, the maximum likelihood estimator (MLE) in the second term
can be found as outlined in
Section~\ref{sec:generalized-likelihood-ratio}.
Figure~\ref{fig:2a} shows the empirical distribution of the resulting MLEs,
compared to the distribution of the true MLEs. It clearly suggests that
the proposed inference procedure yields in this case an unbiased estimator
($\mathbb{E} [\hat \gamma] = 0.0501$, as empirically estimated over $1000$
trials). Note that had the underlying ratio not been accurate enough because of
poor training or poor calibration, the two distributions would not be exactly
overlapping, indicating bias. From this, the initial $\Lambda$ test statistic can finally
be evaluated to determine in this case that the null should not be rejected.  Figure \ref{fig:2b} also confirms
that $-2 \log \Lambda$, as approximated with our method, is asymptotically $\chi^2_1$ distributed,
as should be the true statistic when the null is true.

@@

Given observations ${\cal D}$ represented in Figure~\ref{fig:4},
our goal is to infer the nominal values $\alpha=1$ and $\beta=-1$ that
were used for generating this data.
We use the objective function
\begin{equation}\label{eqn:obj-ex2}
    -2 \log \Lambda(\alpha, \beta ) = -2 \log \frac{p({\cal D} | \alpha, \beta)}{ \sup_{\alpha,\beta} p({\cal D} | \alpha, \beta) }
\end{equation}
that we calculate by exploiting the fact that
\begin{equation}\label{eqn:obj-ex2-expanded}
\log \frac{p({\cal D} | \alpha, \beta)}{ \sup_{\alpha,\beta} p({\cal D} | \alpha, \beta) } = \log \frac{p({\cal D} | \alpha, \beta)}{ p({\cal D} | \alpha=0, \beta=0) } - \log \frac{ \sup_{\alpha,\beta} p({\cal D} | \alpha, \beta) }{ p({\cal D} | \alpha=0, \beta=0)}.
\end{equation}
Following recommendations from Section~\ref{sec:param-clf} to reduce
computational requirements, both terms in the right hand side of Eqn.~\ref{eqn:obj-ex2-expanded} are evaluated using a single
parameterized approximated  likelihood ratio, in this case based on a parameterized
2-layer neural network where the alternate is fixed to $\alpha=0, \beta=0$.
Since the generative model is not expensive, the classifier output is calibrated
on-the-fly with histograms for  every candidate parameter pair $(\alpha,\beta)$.
Minimizing the constructed approximation of the objective with
Bayesian optimization~\citep{brochu2010tutorial} eventually leads to the
estimated posterior mean shown in Figure~\ref{fig:5c}, which is close to the
true objective illustrated in Figure~\ref{fig:5a}. After $50$ iterations, the
estimated minimizer is found at $(\hat \alpha=1.008, \hat \beta=-1.004)$. For
comparison, Figure~\ref{fig:5b} shows the objective value as directly
approximated by the proposed method for a fixed grid of values. Since $\hat s(\mathbf{x;\alpha,\beta})$ is calibrated on-the-fly,
roughness in the contours can be observed, in particular due to non-smooth changes in the resulting histograms as $\alpha$ and $\beta$ vary.
As discussed in Section~\ref{sec:param-calibration}, parameterized calibration
is a solution to this issue, but is left for now as an area for future work.
Overall, this example further illustrates and confirms the ability
of the proposed method for inference, even in the case of multi-dimensional data where
direct approximations $p(\mathbf{x}|\theta_0)$ and $p(\mathbf{x}|\theta_1)$
are often difficult to construct.
